---
title: "Chronic Illness"
output: html_notebook
---

```{r warning=FALSE, message=FALSE}
library(tidyverse)
library(viridis)

md <- src_sqlite("MelbDatathon2017.sqlite", create=FALSE)
src_tbls(md)
```

## What information do we have on patients?

```{r}
patients <- tbl(md,"Patients") %>% collect(n=Inf) %>%
    mutate(
        Patient_ID = as.integer(Patient_ID),
        age=ifelse(year_of_birth==1900,NA,2017-year_of_birth))
```

Number of patients:

```{r}
nrow(patients)
```

Gender information:

```{r}
table(patients$gender)
```

1900 appears to encode missing year of birth. How many patients have 1900 as year of birth?

```{r}
table(patients$year_of_birth == 1900)
```

Where we do have year of birth, how old are the patients?

```{r}
ggplot(patients,aes(x=age)) + geom_bar()
```


## Chronic illness

A table was provided associating certain drugs with chronic illnesses. How many drugs are associated with each illness?

```{r}
drug_illness <- tbl(md,"ChronicIllness") %>%
    transmute(illness=ChronicIllness, Drug_ID=as.integer(MasterProductID)) %>%
    collect(n=Inf) %>%
    mutate(illness=ifelse(illness=="Chronic Obstructive Pulmonary Disease (COPD)","COPD",illness))

drug_illness %>% count(illness)
```

### How many patients have taken a drug associated with a chronic illness?

```{r}
patient_drug <- tbl(md,"Transactions") %>%
    count(Patient_ID, Drug_ID) %>%
    collect(n=Inf) 

patient_illness <- inner_join(ungroup(patient_drug), drug_illness, by="Drug_ID") %>%
    distinct(Patient_ID, illness)
```

Most of the patients in the data set can be classified as having some chronic illness, or at least having taken a drug that is associated with a chronic illness.

```{r}
table(patients$Patient_ID %in% patient_illness$Patient_ID)
```

```{r}
patient_illness %>% count(illness) %>% 
    mutate(illness=reorder(illness,-n)) %>%
    ggplot(aes(x=illness,y=n)) + geom_bar(stat="identity") + 
    coord_flip() + labs(x="",y="Number of patients")
```

### Which illnesses co-occur in patients most often?

```{r}
co_illness <- rename(patient_illness, illness1=illness) %>%
    inner_join(rename(patient_illness, illness2=illness), by="Patient_ID") %>%
    count(illness1,illness2) %>% ungroup()

ggplot(co_illness, aes(x=reorder(illness1,-n), y=reorder(illness2,-n), fill=log10(n))) +    
    geom_tile(color="black") + coord_fixed() +
    scale_fill_viridis() +
    theme_minimal() + theme( axis.text.x=element_text(angle=90,vjust=0.5,hjust=1) ) +
    labs(x="",y="",fill="log10\nnumber of\npatients")
```

### Illness by gender

```{r}
illness_gender <- patient_illness %>%
    inner_join(filter(patients, gender != "U"), by="Patient_ID") %>%
    count(gender, illness) %>% ungroup()

illness_gender %>%
    mutate(
        illness=reorder(illness,-n),
        n=ifelse(gender=="M",-n,n)) %>%
    ggplot(aes(x=illness,y=n,fill=gender)) + geom_bar(stat="identity") + 
    coord_flip() + labs(x="",y="Number of patients",fill="Gender") + ylim(c(-1.75e5,1.75e5))
```

### Illness by age

What is the median age for each illness?

```{r}
illness_age <- patient_illness %>%
    inner_join(filter(patients, !is.na(age)), by="Patient_ID") %>%
    count(age, illness) %>% ungroup()

median_ages <- illness_age %>% 
    group_by(illness) %>% 
    summarize(median_age=median(rep(age,n))) %>% 
    arrange(median_age)

median_ages
```

```{r}
illness_age %>% 
    mutate(illness=factor(illness,median_ages$illness)) %>%
    ggplot(aes(x=age,y=n)) + geom_bar(stat="identity") + 
    facet_grid(illness~.,scales="free_y") + labs(y="Number of patients", x="Age") +
    theme(strip.text.y = element_text(angle=0))
```




