---
title: "The PBS safety net"
output: html_notebook
---

```{r message=F, warning=F}
library(tidyverse)
library(lubridate)
library(viridis)
```

```{r}
md <- src_sqlite("MelbDatathon2017.sqlite", create=FALSE)
# db_create_index(md$con, "Transactions", c("Patient_ID"))

# Use patients with records through 2016
max_id <- 279200

oz_pieces <- read_csv("../australia_map.csv")
oz_geom <- geom_polygon(data=oz_pieces, 
    aes(group=piece,x=long,y=lat), inherit.aes=FALSE, fill="#cccccc")
# ggplot() + oz_geom

stores <- tbl(md,"Stores") %>% collect(n=Inf) %>%
    mutate(Store_ID = as.integer(Store_ID))

drug_illness <- tbl(md,"ChronicIllness") %>%
    transmute(illness=ChronicIllness, Drug_ID=as.integer(MasterProductID)) %>%
    collect(n=Inf) %>%
    mutate(illness=ifelse(illness=="Chronic Obstructive Pulmonary Disease (COPD)","COPD",illness))

drugs <- tbl(md,"Drugs") %>%
    rename(Drug_ID=MasterProductID) %>%
    mutate(Drug_ID=as.integer(Drug_ID)) %>%
    collect(n=Inf) %>%
    left_join(drug_illness, by="Drug_ID")

stores <- tbl(md,"Stores") %>%
    mutate(Store_ID=as.integer(Store_ID)) %>%
    collect(n=Inf)

desc <- function(names) {
    if (length(names) > 3) names <- c(head(names,2),"...")
    paste(names, collapse="\n")
}
postcode_suburb <- read_csv("../postcode_suburb.csv") %>%
    group_by(postcode) %>% summarize(label=desc(suburb))

postcodes <- read_csv("../postcode_centroids.csv") %>%
    select(postcode=POA_CODE, area=SQKM, long_c, lat_c) %>%
    left_join(postcode_suburb,by="postcode")

patients <- tbl(md,"Patients") %>% collect(n=Inf) %>%
    mutate(
        Patient_ID = as.integer(Patient_ID),
        age=ifelse(year_of_birth==1900,NA,2017-year_of_birth))

records <- tbl(md,"Transactions") %>%
    filter(Patient_ID <= max_id) %>%
    distinct(Patient_ID, Drug_ID, Store_ID, Dispense_Week, 
        PatientPrice_Amt, WholeSalePrice_Amt, GovernmentReclaim_Amt) %>%
    collect(n=Inf)

records <- records %>%
    mutate(day=as.integer(ymd(Dispense_Week))) %>%
    left_join(dplyr::select(drugs, Drug_ID, illness), by="Drug_ID") %>%
    left_join(dplyr::select(stores, Store_ID, postcode), by="Store_ID")


df <- records %>%
    mutate(
        Dispense_Week = ymd(Dispense_Week),
        year=year(Dispense_Week)) %>%
    filter(
        !is.na(PatientPrice_Amt), 
        !is.na(WholeSalePrice_Amt),
        year >= 2011,
        year <= 2016) %>%
    group_by(Patient_ID, year) %>%
    mutate(
        zero_paid = (PatientPrice_Amt == 0),
        concession = any(zero_paid) | (
            any(
                PatientPrice_Amt > 0 &
                PatientPrice_Amt < WholeSalePrice_Amt & 
                WholeSalePrice_Amt > 10
            ) & !any(
                PatientPrice_Amt < WholeSalePrice_Amt &
                PatientPrice_Amt > 10
            )
        )
    ) %>%
    ungroup() %>%
    mutate(
        zero_paid=factor(ifelse(zero_paid,"Patient pays nothing","Co-payment"),
            c("Patient pays nothing", "Co-payment")),
        concession=factor(ifelse(concession,"Concession","Non-concession"))
    )
```

## A striking yearly pattern in purchases

Only patients with complete 2016 transactions have been used. I omit data from 2017, as our interest here will be in yearly trends.

```{r}
df %>%
count(Dispense_Week) %>%
ggplot(aes(x=Dispense_Week, y=n)) + geom_col(width=7) +
    labs(x="", y="Number of prescription purchases")
```

## The pattern appears to be due to the PBS safety net

https://www.humanservices.gov.au/customer/services/medicare/pharmaceutical-benefits-scheme-pbs-safety-net

The Pharmaceutical Benefits Scheme is a co-payment system -- the patient pays up to some amount for a medicine and the government pays the rest. However if a patient has paid beyond a certain amount they hit the "safety net" and the government pays the full amount for medications for the rest of the year.

Concession card holders currently pay \$6.30 per purchase but nothing beyond a total of \$378 each year. Concession cards are given to various people including the elderly, veterans, unemployed people, and disabled people.

Non-concession card holders currently pay \$38.80 per purchase but \$6.30 beyond a total of \$1,494.90 each year.

Slightly more may be paid for brands more expensive than the cheapest for a drug.

I've inferred concession card holding, for each patient in each year, if a patient has paid zero for any medication, or if when a patient has purchased medicine with wholesale price more than \$10 and the government has made a contribution the patient has always paid less than \$10.

```{r}
df %>%
#mutate(Dispense_Week = floor_date(Dispense_Week, unit="6 months") %m+% months(3)) %>%
count(Dispense_Week, zero_paid, concession) %>%
ungroup() %>%
ggplot(aes(x=Dispense_Week, y=n, fill=zero_paid)) + geom_col(width=7) +
    facet_grid(concession ~ .) + 
    labs(fill="", x="", y="Number of prescription purchases")
```

Binning by half-year shows the pattern more clearly. When people have to pay, they purchase less medicine.

```{r}
df %>%
mutate(Dispense_Week = floor_date(Dispense_Week, unit="6 months") %m+% months(3)) %>%
count(Dispense_Week, zero_paid, concession) %>%
ungroup() %>%
ggplot(aes(x=Dispense_Week, y=n, fill=zero_paid)) + geom_col(width=100) +
    facet_grid(concession ~ .) + 
    labs(fill="", x="", y="Number of prescription purchases")
```


## Chronic illness purchases are also lower early in the year

Some people seem to be rationing their use of such things as anti-depressants, heart medication, asthma or bronchitis medication, diabetes medication, and medication for high cholesterol.

```{r fig.height=8, fig.width=5}
df %>%
filter(!is.na(illness)) %>%
#mutate(Dispense_Week = floor_date(Dispense_Week, unit="6 months") %m+% months(3)) %>%
count(illness, Dispense_Week, zero_paid, concession) %>%
ungroup() %>%
ggplot(aes(x=Dispense_Week, y=n, fill=zero_paid)) + geom_col(width=7) +
    facet_grid(illness ~ concession, scales="free_y") + 
    labs(fill="", x="", y="Number of prescription purchases") +
    theme(strip.text.y = element_text(angle=0))
```


```{r fig.height=8, fig.width=5}
df %>%
filter(!is.na(illness)) %>%
mutate(Dispense_Week = floor_date(Dispense_Week, unit="6 months") %m+% months(3)) %>%
count(illness, Dispense_Week, zero_paid, concession) %>%
ungroup() %>%
ggplot(aes(x=Dispense_Week, y=n, fill=zero_paid)) + geom_col(width=100) +
    facet_grid(illness ~ concession, scales="free_y") + 
    labs(fill="", x="", y="Number of prescription purchases") +
    theme(strip.text.y = element_text(angle=0))
```

This doesn't seem like an ideal way to manage chronic conditions, so there may be a case for modifying how the PBS safety net works.


## Where are people most often paying zero?

```{r}
p <- df %>%
    group_by(postcode) %>%
    summarize(n=n(), prop_safety_net=mean(zero_paid == "Patient pays nothing")) %>%
    filter(n >= 100) %>%
    inner_join(postcodes, by="postcode") %>%
    ggplot(aes(x=long_c,y=lat_c,size=n,color=prop_safety_net)) + 
        oz_geom + coord_fixed() + 
        geom_point() +
        scale_color_viridis() +
        labs(x="",y="",size="Number of\npurchases", 
            color="Proportion of\nzero payment\npurchases")

p + 
    scale_y_continuous(limits=c(-38.5,-37.5),oob=function(a,b)a) +
    scale_x_continuous(limits=c(144,145.5),oob=function(a,b)a) +
    scale_size_area(max_size=4)
```




